\modLP{Identification of} Biological Regulatory Networks from Process Hitting models

### 2014-11-12 parts-phage-lambda.tex
fix figures
corrections mineures
last minor corrections

@@

We consider here the model with four genes proposed in \cite{thieffry_dynamical_1995},
which will serve as example to detail the inference method proposed in the previous sections.
This model can be straightforwardly represented under the form of a Process Hitting
with 4 components %($\Gamma = \{ \cI, \cro, \cII, \N \}$)
and 4 cooperative sorts: %($\PHs \setminus \Gamma = \{ \cIcrocII, \cIcroN, \cIcro \}$).
\[\PHs = \{ \underbrace{\cI, \cro, \cII, \N}_{\Gamma},
  \underbrace{\cIcrocII, \cIcroN, \cIcro, \crocII}_{\Delta} \}\]
The four components have respectively 3, 4 2 and 2 expression levels,
which we denote:
\begin{align*}
  \PHl_{\cI} &= \{ \cI_0, \cI_1, \cI_2 \} &
  \PHl_{\cro} &= \{ \cro_0, \cro_1, \cro_2, \cro_3 \} \\
  \PHl_{\cII} &= \{ \cII_0, \cII_1 \} &
  \PHl_{\N} &= \{ \N_0, \N_1 \} \\
\end{align*}

@@

\begin{figure}[p]
\centering
\scalebox{1.3}{
\begin{tikzpicture}
  %\path[use as bounding box] (-4,-1.9) rectangle (4.5,3.9);

@@

\begin{figure}[t]
\centering
\scalebox{1.2}{
\begin{tikzpicture}[grn]
  %\path[use as bounding box] (-1.3,-0.75) rectangle (3.5,1.5);
  \node[inner sep=0] (cI) at (0,2.5) {$\cI$};
  \node[inner sep=0] (cro) at (2.5,2.5) {$\cro$};
  \node[inner sep=0] (cII) at (0,0) {$\cII$};
  \node[inner sep=0] (N) at (2.5,0) {$\N$};
  \path[->]
    (cI) edge[bend right] node[elabel, above=-4pt] {$-2$} (cro)
          edge[bend right] node[elabel, left=-2pt] {$-2$} (cII)
          edge[pos=.85] node[elabel, above=-2pt] {$-1$} (N)
    (cro) edge[bend right] node[elabel, above=-4pt] {$-1$} (cI)
          edge[loop right] node[elabel, below=-2pt] {$-3$} (cro)
          edge[pos=.65] node[elabel, below=-2pt] {$-3$} (cII)
          edge node[elabel, right=-4pt] {$-2$} (N)
    (cII) edge[bend right] node[elabel, left=-3pt] {$+1$} (cI)
    (N) edge node[elabel, above=-4pt] {$+1$} (cII);
\end{tikzpicture}
}
\caption{\label{fig:phage-lambda-ig}%
  Result of the IG inference performed on the Process Hitting model
  of the phage lambda immunity response
  taken from \cite{thieffry_dynamical_1995}.
}
\end{figure}

@@

We note two main differences between the parametrization obtained by our inference
and the one proposed in \cite{thieffry_dynamical_1995}:
\begin{itemize}
  \item The absence of the self-regulation of $\cI$ automatically reduces the size
    of the parametrization, but this does not modify the dynamics, as mentioned previously;
  \item Our inference returns $K_{\cro, \{\cI;\cro\}} = 2$, \modLP{which is a value
    equivalent to the original $0$ value according to \pref{rem:K-equiv}:
    the resulting dynamics is the same, and $1$ would also have been an
    acceptable value}.
\end{itemize}

>>>>

% vi:spell spelllang=en:

@@

We consider here the model with four genes proposed in \cite{thieffry_dynamical_1995},
which will serve as example to detail the inference method proposed in the previous sections.
This model can be straightforwardly represented under the form of a Process Hitting
with 4 components %($\Gamma = \{ \cI, \cro, \cII, \N \}$)
and 4 cooperative sorts: %($\PHs \setminus \Gamma = \{ \cIcrocII, \cIcroN, \cIcro \}$).
\[\PHs = \{ \underbrace{\cI, \cro, \cII, \N}_{\Gamma},
  \underbrace{\cIcrocII, \cIcroN, \cIcro, \crocII}_{\Delta} \}\]
The four components have respectively 3, 4, 2 and 2 expression levels,
which we denote:
\begin{align*}
  \PHl_{\cI} &= \{ \cI_0, \cI_1, \cI_2 \} &
  \PHl_{\cro} &= \{ \cro_0, \cro_1, \cro_2, \cro_3 \} \\
  \PHl_{\cII} &= \{ \cII_0, \cII_1 \} &
  \PHl_{\N} &= \{ \N_0, \N_1 \} \\
\end{align*}

@@

\begin{figure}
\centering
\scalebox{1.3}{
\begin{tikzpicture}
  %\path[use as bounding box] (-4,-1.9) rectangle (4.5,3.9);

@@

\begin{figure}
\centering
\scalebox{1.2}{
\begin{tikzpicture}[grn]
  %\path[use as bounding box] (-1.3,-0.75) rectangle (3.5,1.5);
  \node[inner sep=0] (cI) at (0,2.5) {$\cI$};
  \node[inner sep=0] (cro) at (2.5,2.5) {$\cro$};
  \node[inner sep=0] (cII) at (0,0) {$\cII$};
  \node[inner sep=0] (N) at (2.5,0) {$\N$};
  \path[->]
    (cI) edge[bend right] node[elabel, above=-4pt] {$-2$} (cro)
          edge[bend right] node[elabel, left=-2pt] {$-2$} (cII)
          edge[pos=.85] node[elabel, above=-2pt] {$-1$} (N)
    (cro) edge[bend right] node[elabel, above=-4pt] {$-1$} (cI)
          edge[loop right] node[elabel, below=-2pt] {$-3$} (cro)
          edge[pos=.65] node[elabel, below=-2pt] {$-3$} (cII)
          edge node[elabel, right=-4pt] {$-2$} (N)
    (cII) edge[bend right] node[elabel, left=-3pt] {$+1$} (cI)
    (N) edge node[elabel, above=-4pt] {$+1$} (cII);
\end{tikzpicture}
}
\caption{\label{fig:phage-lambda-ig}%
  Result of the IG inference performed on the Process Hitting model
  of the phage lambda immunity response
  taken from \cite{thieffry_dynamical_1995}.
}
\end{figure}

@@

We note two main differences between the parametrization obtained by our inference
and the one proposed in \cite{thieffry_dynamical_1995}:
\begin{itemize}
  \item The absence of the self-regulation of $\cI$ automatically reduces the size
    of the parametrization, but this does not modify the dynamics, as mentioned previously;
  \item Our inference returns $K_{\cro, \{\cI;\cro\}} = 2$, \modLP{which is a value
    equivalent to the original value $0$ according to \pref{rem:K-equiv}:
    the resulting dynamics is the same, and $1$ would also have been an
    acceptable value}.
\end{itemize}

### 2014-11-12 parts-tr2global.tex
last minor corrections

Process Hitting (PH) and Thomas modelling rely on two different paradigms for the specification of
network dynamics:
Thomas parametrization defines a global function
which associates to each component and each resource configuration a level toward which the component
eventually evolves in the given configuration.
Such a function is always deterministic (whereas the associated asynchronous dynamics can be
indeterministic due to concurrent evolution of components).
In contrast, 
PH models specify a list of local transitions between the levels (processes) of the components (sorts)
that are conditioned with the presence of other processes.
In addition, PH allows the use of intermediate sorts, such as cooperative sorts, that do not refer
to the components of the network.

Hence, identifying Thomas models from PH requires to lump a set of transitions that can be
applied in a given configuration into a single process that correspond to the level toward which the
component evolve.
This step is detailed in \pref{ssec:focal} in which is defined the $\focals$ function associating 
the farthest reachable processes, called \emph{focal processes}, of a given sort
for a given configuration.
It is possible that several different focal processes are identified, indicating an indeterministic
evolution of the sort;
and it is also possible that no focal process exist, when in presence of terminal cycles of
transitions.
In those cases, there is no possible Thomas specification for the PH.
The \pref{ssec:dfocals} discusses on constraints for obtaining deterministic
(local) behaviours in PH.

>>>>

Process Hitting (PH) and Thomas modelling rely on two different paradigms for the specification of
network dynamics:
Thomas parametrization defines a global function
which associates to each component and each resource configuration a level toward which the component
eventually evolves in the given configuration.
Such a function is always deterministic (whereas the associated asynchronous dynamics can be
nondeterministic due to concurrent evolution of components).
In contrast, 
PH models specify a list of local transitions between the levels (processes) of the components (sorts)
that are conditioned with the presence of other processes.
In addition, PH allows the use of intermediate sorts, such as cooperative sorts, that do not refer
to the components of the network.

Hence, identifying Thomas models from PH requires to lump a set of transitions that can be
applied in a given configuration into a single process that correspond to the level toward which the
component evolves.
This step is detailed in \pref{ssec:focal} in which is defined the $\focals$ function associating 
the farthest reachable processes, called \emph{focal processes}, of a given sort
for a given configuration.
It is possible that several different focal processes are identified, indicating an nondeterministic
evolution of the sort;
and it is also possible that no focal process exist, when in presence of terminal cycles of
transitions.
In those cases, there is no possible Thomas specification for the PH.
The \pref{ssec:dfocals} discusses on constraints for obtaining deterministic
(local) behaviours in PH.
